Experience and prey behaviour jointly shape individual foraging specialisation and success in a predator-prey videogame

### 2023-01-26 manuscript/Chapter2-manuscript.Rmd
added versions of MS

---
title: "Hunting experience and prey variability jointly shape individual foraging specialisation in a predator-prey videogame"
output: 
  word_document:
    reference_docx: Chapter2-styles.docx
    fig_caption: yes
    pandoc_args: ["-Fpandoc-crossref"] # to have eq. numbers
bibliography: references.bib
csl: e-letters.csl
---

@@

For each level of experience (i.e. novice, intermediate, advanced), we estimated the mean speed of individual predators, and a heterogeneous structure of the residuals such that we could estimate their intra-individual variance (IIV) in speed (i.e. specialization). We followed the same structure for the preys' speed to estimate among individual differences in the mean and variance in speed of the prey encountered by predators. For hunting success, we only modeled the mean part of the equation to estimate among individual differences in mean hunting success. For all variables, we controlled for the duration of the match standardized in mean and unit variance. The model has the following form, with each equation repeated three times (i.e. one for each level of experience) :

@@

where $env$, $av$, and $id$ are the vectors of random environment, random avatar, and random individual identity effects associated with their incidence matrices $M$, $N$, and $Z$. $b$ is the vector of fixed effects with its incidence matrix $X$. The $m$ and $d$ subscripts indicate that an estimate is from the mean or dispersion part of the model, respectively. We assumed that all random effects followed a Gaussian distribution with estimated variance ($env \sim N(0,\ I\ \sigma^2_{env})$, $av \sim N(0,\ I\ \sigma^2_{av})$, and $id \sim N(0,\ I\ \sigma^2_{id})$). The variance parameters $\sigma^2_{env}$, $\sigma^2_{av}$, and $\sigma^2_{id}$, correspond to the among environment, among avatar, and among individual variances, respectively. $I$ is the identity matrix, with $I\sigma^2$ indicating that the random effects are independently and identically distributed. Here, we assume that the residuals vary among individuals where $e \sim N(0,\ Diag\ [\sigma^2_{e}])$, $[\frac{id_m}{id_v}] \sim N\left(\begin{array}{cc} 0 \\ \\ 0 \end{array},\ \Sigma_{id} \otimes\ I \right)$, and $\Sigma_{id}$ is a variance covariance matrix of the individual identity random effect. Equations (1) and (2) were parametrized with a Gaussian distribution family (predator and prey speed), while equation (3) was parametrized with a beta-binomial distribution family (predator hunting success). We parametrized the model to run four MCMC chains with 1000 posterior samples for each parameter. To do so, we ran 2500 iterations with a thinning set to 8, with the first 500 iterations used as warmups. We explain how we specified the priors of the MDHGLM in the supporting information.

We estimated the correlation between the individual random effect on the mean and on the dispersion within each trait (e.g. $cor(\sigma^2_{id_{m}\ pred}\ \sigma^2_{id_{d}\ pred})$) and among each trait (e.g. $cor(\sigma^2_{id_{m}\ pred}\ \sigma^2_{id_{d}\ prey})$) within a level of experience. In the first case, this allows us to assess, for example, whether predators that are on average slower/faster are more specialized or flexible. In the second case, the correlations enables us to estimate, for example, whether slower/faster (or more specialized/flexible) predators encountered prey that were faster/slower, or if slower/faster predators had lower/greater hunting success. We also evaluated these correlations among the different levels of experience (e.g. $cor(\sigma^2_{id_{m}\ pred\ novice}\ \sigma^2_{id_{d}\ pred\ intermediate})$). The latter "character-state" approach [@Via.Lande1985;@viaAdaptivePhenotypicPlasticity1995], enables us to investigate whether predators express similar behaviours across each developmental stage. The correlations were estimated assuming a multivariate Gaussian distribution.

Lastly we calculated the coefficient of variation in the mean (termed $CV_I$) and dispersion (termed $CV_p$) part of the predator's speed for each stage of experience, allowing us to report standardized effect sizes of individual variance for comparison across traits and studies [@cleasbyQuantifyingPredictabilityBehaviour2015;@Holtmann.etal2017;@Dingemanse.etal2022]. This also allowed us to evaluate whether predators displayed increasing/decreasing differences in specialization at each stage of experience.

@@

Our results confirm that individual predators displayed important differences in the development of their expertise, as the best model was the one with group-level smoothers exclusively (Table I). For instance, with experience, some predators steadily increased their hunting success while others displayed a decrease in prey capture (Figure 1A). Otherwise, there were players that maintained either a high or low success across their experience (Figure 1A). Another pattern that emerged was that some individuals optimized their success at an experience level that was below what we considered as "advanced" (i.e. >300 matches played).

@@

Contrary to our predictions, we detected only minimal differences in movement speed among individual predators, and these differences did not change with experience (Figure 2A). However, individuals differed in their degree of flexibility, and these differences increased marginally with experience (Figure 2B). Overall, individual predators encountered prey with similar average speeds (Figure 2A), and our results do not support our prediction that predators would encounter prey with varying levels of speed (Figure 2B).

```{r fig2, out.width = "100%", fig.cap = "**Figure 2.** Coefficient of variation for the predator's speed and for the speed of the prey encountered at each level of experience. (A) Coefficient of variation of the among individual differences in average behaviour. (B) Coefficient fo variation of the among individual differences in intra individual variance (i.e. specialization)."}

@@

We predicted that, with experience, predators would become either increasingly flexible, or, more specialized. Here, we found that predators became generally more flexible with experience, but this increase in flexibility was only minimal (Figure 3 A and B). Thus we cannot conclude that experience had an important effect on behavioural specialization. Yet, as we have outlined in Figure 2, it appears that among individual differences in specialization increase when individuals are advanced compared to when they are novice (Figure 3 A and B).

```{r fig3, out.width = "100%", fig.cap = "**Figure 3.** Distribution of the intra individual behavioural variation (i.e. specialization) of predators when they were (A) novices and then (B) advanced hunters. The red band displays the 95% CI of the intercept for the intra individual standard deviation in predator speed."}

@@

 # Extract te draws
 draws <- data.table(
     as_draws_df(
         fit,
         variable = c("cor"),
         regex = TRUE)
 )

@@

At all experience levels, predators that were on average faster were also travelling faster in all their matches ($cor_{novice}$ = `r toString(format(cor_tab[1,2]))` [`r toString(format(cor_tab[1,3]))`, `r toString(format(cor_tab[1,4]))`], $cor_{interm}$ = `r toString(format(cor_tab[6,2]))` [`r toString(format(cor_tab[6,3]))`, `r toString(format(cor_tab[6,4]))`], $cor_{adv}$ = `r toString(format(cor_tab[15,2]))` [`r toString(format(cor_tab[15,3]))`, `r toString(format(cor_tab[15,4]))`]) (i.e. they were more specialized at travelling fast). Thus, slower predators were also more flexible in their movement through time (Figure 4). This pattern became stronger as predators gained experience (Figure 4). In general, predators tended to be faster when they encountered prey that were on average faster ($cor_{novice}$, $cor_{interm}$, $cor_{adv}$) (Figure 4). They were also more specialized at being fast when prey were on average faster (Figure 4).

Moreover, predators favored and specialized in a faster foraging mode when encounters with prey were more predictable, and these relationships also became stronger with experience (Figure 4).

The foraging strategies of predators, that is, the correlations between the mean and variance of speed, remained structurally stable among the levels of experience. For instance, novice predators that tended to be faster also kept this strategy as they gained experience (Figure 4 inside the black squares). Overall, we detected the same patterns for within-experience state correlations. The correlations between trait values were also stronger among novice-intermediate and intermediate-advanced states than between novice-advanced states, indicating that behavioural changes in predators occurred as they gained experience.

```{r fig4, out.width = "100%", fig.cap = "**Figure 4.** Full correlation matrix of the MDHGLM relating the mean and IIV speed of the predators, the mean and IIV in speed of the prey they encountered, and mean hunting success. Starting from the upper part of the plot are displayed the five parameters for novice predators (gray), the five for intermediate predators (yellow), and the five for advanced predators (blue). Larger dark blue circles indicate stronger positive correlations, while larger dark red circles indicate stronger negative correlations. The black squares indicate correlations between different experience levels (i.e. character states)."}

knitr::include_graphics(here::here("outputs", "05_outputs_figures", "05_figure4.png"))

@@

We did not detect significant relationships between the predators' average nor intra individual variation in speed and hunting success (Figure 4), indicating that both slow/flexible and fast/specialist predators were both equally successful. However, when predators were novice, there was a strong positive correlation between the variation in prey that they encountered and their hunting success. Hence, novice individuals that experienced more unpredictable prey encounters at this stage were also more successful. This relationship was decoupled at the later experience levels, such that unpredictable encounters at intermediate and advanced stages were not associated with hunting success anymore.

>>>>

---
title: "Experience and prey behaviour jointly shape individual foraging specialisation and success in a predator-prey videogame"
output: 
  word_document:
    reference_docx: Chapter2-styles.docx
    fig_caption: yes
    pandoc_args: ["-Fpandoc-crossref"] # to have eq. numbers
bibliography: references.bib
csl: e-letters.csl
---

@@

For each level of experience (i.e. novice, intermediate, advanced), we estimated the mean speed of individual predators, and a heterogeneous structure of the residuals such that we could estimate their intra-individual variation (IIV) in speed (i.e. specialization/flexibility). We followed the same structure for the preys' speed to estimate among individual differences in the mean and variance in speed of the prey encountered by predators. For hunting success, we only modeled the mean part of the equation to estimate among individual differences in mean hunting success. For the predator and the prey's speed, we controlled for the average rank of the prey that they encountered. The ranking system in *DBD* was implemented by the company to pair players in a match based on their skill (https://deadbydaylight.fandom.com/wiki/Rank). We used the mean rank of the four prey and applied a square-root transformation. For the predator's hunting success, we controlled for the square-root of the duration of the match. The model has the following form, with each equation repeated three times (i.e. one for each level of experience) :

@@

where $env$, $av$, and $id$ are the vectors of random environment, random avatar, and random individual identity effects associated with their incidence matrices $M$, $N$, and $Z$. $b$ is the vector of fixed effects with its incidence matrix $X$. The $m$ and $d$ subscripts indicate that an estimate is from the mean or dispersion part of the model, respectively. We assumed that all random effects followed a Gaussian distribution with estimated variance ($env \sim N(0,\ I\ \sigma^2_{env})$, $av \sim N(0,\ I\ \sigma^2_{av})$, and $id \sim N(0,\ I\ \sigma^2_{id})$). The variance parameters $\sigma^2_{env}$, $\sigma^2_{av}$, and $\sigma^2_{id}$, correspond to the among environment, among avatar, and among individual variances, respectively. $I$ is the identity matrix, with $I\sigma^2$ indicating that the random effects are independently and identically distributed. Here, we assume that the residuals vary among individuals where $e \sim N(0,\ Diag\ [\sigma^2_{e}])$, $[\begin{array}{cc} id_m \\ id_v \end{array}] \sim N\left(\begin{array}{cc} 0 \\ \\ 0 \end{array},\ \Sigma_{id} \otimes\ I \right)$, and $\Sigma_{id}$ is a variance covariance matrix of the individual identity random effect. Equations (1) and (2) were parametrized with a Gaussian distribution family (predator and prey speed), while equation (3) was parametrized with a beta-binomial distribution family (predator hunting success). We parametrized the model to run four MCMC chains with 1000 posterior samples for each parameter. To do so, we ran 2500 iterations with a thinning set to 8, with the first 500 iterations used as warmups. We explain how we specified the priors of the MDHGLM in the supporting information.

We estimated the correlation between the individual random effect on the mean and on the dispersion within each trait (e.g. $cor(\sigma^2_{id_{m}\ pred}\ \sigma^2_{id_{d}\ pred})$) and among each trait (e.g. $cor(\sigma^2_{id_{m}\ pred}\ \sigma^2_{id_{d}\ prey})$) within a level of experience. In the first case, this allows us to assess, for example, whether predators that are on average slower/faster are more specialized or flexible. In the second case, the correlations enables us to estimate, for example, whether predators that encountered prey that were faster/slower adopted a slower/faster (or more specialized/flexible) strategy, or if slower/faster predators had lower/greater hunting success. We also evaluated these correlations among the different levels of experience (e.g. $cor(\sigma^2_{id_{m}\ pred\ novice}\ \sigma^2_{id_{d}\ pred\ intermediate})$). The latter "character-state" approach [@Via.Lande1985;@viaAdaptivePhenotypicPlasticity1995], enables us to investigate whether predators express similar behaviours across each developmental stage. The correlations were estimated assuming a multivariate Gaussian distribution.

Lastly we calculated the intra-class correlation coefficients ($ICC$, i.e. repeatability) of each random effect on the mean part of the predator's speed and prey speed to estimate the percentage of variation in speed attributed to differences among environments, avatars, and individuals. We also report the coefficient of variation in the mean (termed $CV_I$) and dispersion (termed $CV_p$) part of the predator and prey's speed, at each stage of experience, to report standardized effect sizes of individual variance for comparison across traits and studies [@cleasbyQuantifyingPredictabilityBehaviour2015;@Holtmann.etal2017;@Dingemanse.etal2022]. This also allowed us to evaluate whether predators displayed increasing/decreasing differences in specialization at each stage of experience.

@@

Our results confirm that individual predators displayed important differences in the development of their expertise, as the best model was the one with group-level smoothers exclusively (Table I). For instance, with experience, some predators steadily increased their hunting success while others displayed a decrease in prey capture (Figure 1A-S1A-S1B). Otherwise, there were players that maintained either a high or low success across their experience (Figure 1A). Another pattern that emerged was that some individuals optimized their success at an experience level that was below what we considered as "advanced" (i.e. >300 matches played).

@@

We detected among individual differences in the average movement speed of predators ($ICC\ predator\ ID$, Table 2). These differences decreased at the intermediate stage, and increased at the advanced stage (Table 2). Among-individual differences in mean predator speed was slightly higher at the advanced stage compared to the novice stage (Table 2). However, the highest amount of variation in mean predator speed was due to differences among predator avatars, and differences increased with experience ($ICC\ avatar\ ID$, Table 2). We did not detect variation in mean predator speed among the virtual environments ($ICC\ environment\ ID$, Table 2). Individual predators displayed considerable differences in their degree of specialization, and these differences increased with experience ($CV_p\ predator\ ID$, Table 2). Thus, the predator population is composed of individual specialists and generalists in their movement speed, and individuals further distinguish themselves in these strategies as they gain experience (Figure 2).

We predicted that, with experience, predators would become either increasingly flexible, or, more specialized. Our results do not support our prediction, as the pattern was more complex. Indeed, we found that predators became more flexible with experience at the intermediate stage, but when they became advanced hunters, they returned to a level of flexibility similar to when they were novices ($Intercept$, Table 2, Figure 2). Thus, instead of increasing linearly, our results suggest that behavioural specialization changed in a nonlinear (quadratic) manner with experience.

```{r table2, ft.align = "left"}

tab2 <- readRDS(here::here("outputs", "05_outputs_tables", "05_table2.rds"))

tab2 <- autofit(tab2)

tab2 <- tab2 %>% 
  footnote(
    i = 1, j = 1,
    part = "header",
    value = as_paragraph(
      "* We applied the exponent to the intercept values of sigma parameters, estimated on a log scale, and then squared them, such that values are mean posterior residual variances. We back-transformed the hunting success values, estimated on a logit scale, back to a probability scale."),
    ref_symbols = " "
  ) %>%
  font(fontname = "Times New Roman", part = "all") %>%
  width(j = c(1:5), width = 3.7, unit = "cm")

set_caption(
  tab2,
  caption = "Posterior mean estimates and 95% credible intervals computed by the multivariate double-hierarchical generalized linear model of predator speed, prey speed, and predator hunting success.",
  style = "Table Caption"
)

As we predicted, individual predators encountered prey with varying levels of speed ($ICC\ predator\ ID$, Table 2). Similar to the predator's speed, prey encounters varied more among predators at the novice and advanced levels, and predators differed more in the mean speed of prey that they encountered at the advanced stage (Table 2). Thus, some individuals encountered prey that were on average slower, while others encountered prey that where on average faster. These encounters also varied slightly among the predator avatars and virtual environments, but did not change with experience (Table 2). However, predators differed only weakly in the variability between prey encounters that they experienced ($CV_p\ predator\ ID$, Table 2). Hence, while predators did differ in the mean speed of the prey that they encountered, they all experienced similar degrees of variation among encounters (Table 2).

```{r fig2, out.width = "100%", fig.cap = "**Figure 2.** Distribution of the intra individual behavioural variation (i.e. specialization) of predators when they were (A) novices and then (B) advanced hunters. The red band displays the 95% CI of the intercept for the intra individual standard deviation in predator speed."}

@@

@@

 # Extract the draws
 draws <- data.table(
     as_draws_df(
         fit,
         variable = c("cor"),
         regex = TRUE)
 )

@@

At all experience levels, predators that were on average faster (mean) were also travelling faster in all their matches (IIV) ($cor_{novice}$ = `r toString(format(cor_tab[1,2]))` [`r toString(format(cor_tab[1,3]))`, `r toString(format(cor_tab[1,4]))`], $cor_{interm}$ = `r toString(format(cor_tab[6,2]))` [`r toString(format(cor_tab[6,3]))`, `r toString(format(cor_tab[6,4]))`], $cor_{adv}$ = `r toString(format(cor_tab[15,2]))` [`r toString(format(cor_tab[15,3]))`, `r toString(format(cor_tab[15,4]))`]) (i.e. they were more specialized at travelling fast). Thus, slower predators were also more flexible in their movement through time (Figure 3). In general, predators tended to be faster (mean) when they encountered prey that were on average faster (mean) ($cor_{novice}$ = `r toString(format(cor_tab[16,2]))` [`r toString(format(cor_tab[16,3]))`, `r toString(format(cor_tab[16,4]))`], $cor_{interm}$ = `r toString(format(cor_tab[31,2]))` [`r toString(format(cor_tab[31,3]))`, `r toString(format(cor_tab[31,4]))`], $cor_{adv}$ = `r toString(format(cor_tab[50,2]))` [`r toString(format(cor_tab[50,3]))`, `r toString(format(cor_tab[50,4]))`]) (Figure 4). At the three experience levels, there were only weak correlations between predator IIV and mean prey speed ($cor_{novice}$ = `r toString(format(cor_tab[17,2]))` [`r toString(format(cor_tab[17,3]))`, `r toString(format(cor_tab[17,4]))`], $cor_{interm}$ = `r toString(format(cor_tab[32,2]))` [`r toString(format(cor_tab[32,3]))`, `r toString(format(cor_tab[32,4]))`], $cor_{adv}$ = `r toString(format(cor_tab[51,2]))` [`r toString(format(cor_tab[51,3]))`, `r toString(format(cor_tab[51,4]))`]) (Figure 3). Similarly, predators specialized in a faster foraging mode (IIV) when encounters with prey were more predictable (IIV) ($cor_{novice}$ = `r toString(format(cor_tab[23,2]))` [`r toString(format(cor_tab[23,3]))`, `r toString(format(cor_tab[23,4]))`], $cor_{interm}$ = `r toString(format(cor_tab[40,2]))` [`r toString(format(cor_tab[40,3]))`, `r toString(format(cor_tab[40,4]))`], $cor_{adv}$ = `r toString(format(cor_tab[61,2]))` [`r toString(format(cor_tab[61,3]))`, `r toString(format(cor_tab[61,4]))`]), but the correlations were weak (Figure 3). In the four cases described above, the correlations became stronger as predators gained experience (Figure 3).

The foraging strategies of predators, that is, the correlations between the mean and IIV of speed, remained structurally stable among the levels of experience. For instance, novice predators that tended to be faster also kept this strategy as they gained experience (Figure 3 inside the black squares). Overall, we detected the same patterns for within-experience state correlations. The correlations between trait values were also stronger among novice-intermediate and intermediate-advanced states than between novice-advanced states, indicating that behavioural changes in predators occurred as they gained experience.

```{r fig3, out.width = "100%", fig.cap = "**Figure 3.** Full correlation matrix of the MDHGLM relating the mean and IIV speed of the predators, the mean and IIV in speed of the prey they encountered, and mean hunting success. Starting from the upper part of the plot are displayed the five parameters for novice predators (gray), the five for intermediate predators (yellow), and the five for advanced predators (blue). Larger dark blue circles indicate stronger positive correlations, while larger dark red circles indicate stronger negative correlations. The black squares indicate correlations between different experience levels (i.e. character states)."}

@@

We did not detect strong correlations between the predators' average speed and hunting success ($cor_{novice}$ = `r toString(format(cor_tab[67,2]))` [`r toString(format(cor_tab[67,3]))`, `r toString(format(cor_tab[67,4]))`], $cor_{interm}$ = `r toString(format(cor_tab[81,2]))` [`r toString(format(cor_tab[81,3]))`, `r toString(format(cor_tab[81,4]))`], $cor_{adv}$ = `r toString(format(cor_tab[96,2]))` [`r toString(format(cor_tab[96,3]))`, `r toString(format(cor_tab[96,4]))`]), nor between the IIV in speed and hunting success ($cor_{novice}$ = `r toString(format(cor_tab[68,2]))` [`r toString(format(cor_tab[68,3]))`, `r toString(format(cor_tab[68,4]))`], $cor_{interm}$ = `r toString(format(cor_tab[82,2]))` [`r toString(format(cor_tab[82,3]))`, `r toString(format(cor_tab[82,4]))`], $cor_{adv}$ = `r toString(format(cor_tab[97,2]))` [`r toString(format(cor_tab[97,3]))`, `r toString(format(cor_tab[97,4]))`]), suggesting that both slower/flexible and faster/specialist predators were both equally successful (Figure 3). However, when predators were novice, there was a strong negative correlation between the average speed of the prey that they encountered and their hunting success ($cor_{novice}$ = `r toString(format(cor_tab[73,2]))` [`r toString(format(cor_tab[73,3]))`, `r toString(format(cor_tab[68,4]))`]). Hence, novice individuals that encountered faster prey were less successful. This relationship was decoupled at the later experience levels, such that encounters with faster prey at intermediate and advanced stages were not as strongly correlated with hunting success anymore ($cor_{interm}$ = `r toString(format(cor_tab[87,2]))` [`r toString(format(cor_tab[87,3]))`, `r toString(format(cor_tab[87,4]))`], $cor_{adv}$ = `r toString(format(cor_tab[102,2]))` [`r toString(format(cor_tab[102,3]))`, `r toString(format(cor_tab[102,4]))`]).

# Discussion
